---
title: "MAP miRNA - Faith"
author: "Based on DESeq2 manuel"
date: "2024/05/11"
output:
  html_document:
    toc: yes
    number_sections: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---
#clear workspace
rm(list= ls())

References:

https://bioconductor.org/packages/release/bioc/html/DESeq2.html
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
https://bioconductor.org/packages/release/bioc/manuals/DESeq2/man/DESeq2.pdf
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.R

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Librairies
```{r librairies, message=FALSE}
# Date
format(Sys.Date(), format="%d %B %Y")
# Language
Sys.setenv(LANG = "en_US.UTF-8")


r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

install.packages("BiocManager")
BiocManager::install("DESeq2")
BiocManager::install("edgeR")
install.packages("kableExtra")
BiocManager::install("apeglm")
BiocManager::install("IHW")
BiocManager::install("vsn")
BiocManager::install("pheatmap")
install.packages("hexbin")
install.packages("rmarkdown")
install.packages("mvtnorm")  


#Libraries
library(DESeq2)
library(edgeR)
library(kableExtra)
library(apeglm)
library("IHW")
library("vsn")
library("pheatmap")
library("RColorBrewer")
library("ggplot2")
library("hexbin")
library("BiocParallel")
library(knitr)

register(SnowParam(4))
```


# Standard workflow
## Input data
### Count matrix input
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input
```{r inputs}
###############
# Count intputs
cts_snoRNA <- (read.csv("./snoRNA_SC/SNORNA_counts.csv", 
                        header = TRUE, 
                        stringsAsFactors = FALSE, 
                        row.names = 1,
                        sep = ","))

# Transposition and matrix conversions.
#cts_snoRNA <- as.matrix(t(cts_snoRNA)) 


###filter to remove S. aureus

cts_snoRNA <- cts_snoRNA[,-c(2:5,7:8,9:10,12:13,14:16,18:20,22,24:25)]


kable(head(cts_snoRNA[, 1:7]))  %>% kable_styling()

dim(cts_snoRNA)
str(cts_snoRNA)




###############
# coldata
# Is factor required for Parity and DIM? If not, Parity and DIM will be as quantitative value.
coldata <- read.csv(file = "./coldata.csv", 
                    sep = "\t", 
                    header = TRUE, 
                    stringsAsFactors = TRUE,  
                    row.names = 1)
coldata <- coldata[order(rownames(coldata)), ] # coldata is ordered by rownames.

coldata$DIM <- as.factor(coldata$DIM)


coldata <- coldata[-c(2:5,7:8,9:10,12:13,14:16,18:20,22,24:25), ]



dim(coldata)
str(coldata)
kable(coldata) %>% kable_styling()



###############
# Names
# It is absolutely critical that the columns of the count matrix and the rows of the coldata are in the same order.

v1 <- rownames(coldata)

cts_snoRNA <- cts_snoRNA[, v1]


dim(cts_snoRNA)
str(cts_snoRNA)


kable(head(cts_snoRNA[, 1:7])) %>% kable_styling()

  


###############
# Double check names
# The following 4 codes must give TRUE!
all(rownames(coldata) %in% colnames(cts_snoRNA)) # Must be TRUE
all(rownames(coldata) == colnames(cts_snoRNA)) # Must be TRUE 
```
#Count Deseqdataset

```{r}
#Mature
dds1 <- DESeqDataSetFromMatrix(countData = cts_snoRNA,
                              colData = coldata,
                              design = ~Lactation_stage + Treatment)


```
### Pre-filtering
```{r Filtering}
#Mature
keep <- rowSums(counts(dds1)) >= 10
dds1 <- dds1[keep,]

```

## Differential expression analysis
```{r DE}
#Mature dds
dds1 <- DESeq(dds1)
resultsNames(dds1)


#Mature res
res1 <- results(dds1, name="Treatment_S.chromogenes_vs_Control")
kable(head(res1)) %>% kable_styling()
kable(summary(res1)) %>% kable_styling()
```
#####Data normalization
```{r}
#### Extracting transformed values

##### Extracting transformed values


# Data transformations and visualization
#{r Transformations}
##### Extracting transformed values

#Mature
vsd1 <- varianceStabilizingTransformation(dds1)
kable(head(assay(vsd1), 3)) %>% kable_styling()
jpeg(filename = "dds_normal_transform.jpg")
meanSdPlot(assay(vsd1))
dev.off()


rld1 <- rlog(dds1, blind=FALSE)
kable(head(assay(vsd1), 3)) %>% kable_styling()
jpeg(filename = "dds_normal_transform.jpg")
meanSdPlot(assay(rld1))
dev.off()

##Effects of transformations on the variance
# this gives log2(n + 1)

#Mature
ntd1 <- normTransform(dds1)
jpeg(filename = "dds_normal_transform.jpg")
meanSdPlot(assay(ntd1))
dev.off()

meanSdPlot(assay(ntd1))

meanSdPlot(assay(vsd1))

meanSdPlot(assay(rld1))

par( mfrow = c( 1, 2 ) )
dds1 <- estimateSizeFactors(dds1)
plot(log2(counts(dds1, normalized=TRUE)[,1:2] + 1),
      pch=16, cex=0.3)
plot(assay(rld1)[,1:2],
      pch=16, cex=0.3)
```


###plot PCA 
```{r}
## Principal component plot of the samples
plotPCA(vsd1,intgroup=c("Treatment"))

#It is also possible to customize the PCA plot using the ggplot function
pcaData <- plotPCA(vsd1, intgroup=c("Treatment"), returnData=TRUE) #intgroup=c("Treatment", "type")
percentVar <- round(100 * attr(pcaData, "percentVar"))
pca <- ggplot(pcaData, aes(PC1, PC2, color=Treatment)) +  #add ", shape=type" in aes()
  geom_point(size=3) +
  geom_text(aes(label= name), vjust = 1)+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
pca
ggsave('PCA_samples_SC.jpg', plot = pca)
```
## Differential expression analysis
```{r DE}
#Mature dds
dds1 <- DESeq(dds1)
resultsNames(dds1)


#Mature res
res1 <- results(dds1, name="Treatment_S.chromogenes_vs_Control")
kable(head(res1)) %>% kable_styling()
kable(summary(res1)) %>% kable_styling()
```


```{r DE}
#res1 <- results(dds1, lfcThreshold = 0.01)
res1 <- results(dds1)
```


### Log fold change shrinkage for visualization and ranking
```{r Log}
resultsNames(dds1)


# Maure resLFC

resLFC1 <- lfcShrink(dds1, coef="Treatment_S.chromogenes_vs_Control")
kable(head(resLFC1)) %>% kable_styling()

#Exporting only the results which pass an adjusted p value threshold

resSig1 <- subset(res1, padj < 0.01)
kable(head(resSig1)) %>% kable_styling()

```


### p-values and adjusted p-values
```{r pvalues}
# order results table by the smallest p value:
resOrdered1 <- res1[order(res1$pvalue),]
summary(resOrdered1)

sum(res1$padj < 0.1, na.rm=TRUE)
sum(res1$pvalue < 0.05, na.rm = TRUE) 
sum(res1$padj < 0.05, na.rm=TRUE) 


## check DE genes with |logFC|>1 & padj < 0.05

#Mature
res0.05 <- results(dds1, lfcThreshold = 1, alpha = 0.05)
summary(res0.05)

```

### Independent hypothesis weighting
```{r weighting}
#Mature
resIHW1 <- results(dds1, filterFun=ihw)
summary(resIHW1)
sum(resIHW1$padj < 0.05, na.rm=TRUE) 
metadata(resIHW1)$ihwResult
```



## Exploring and exporting results
```{r Exploring}
#Dispersion plot

#Mature
plotDispEsts(dds1, ylim = c(1e-6, 1e2))

#plotMA with only y limit
DESeq2::plotMA(res1, ylim=c(-5, 5))


DESeq2::plotMA(resLFC1, ylim=c(-5, 5))


#plotMA with x and y limit

plotMA(res1, ylim=c(-10, 10), xlim = c(0, 6))

plotMA(resLFC1, ylim=c(-10, 10), xlim = c(0, 6))



#After calling plotMA, one can use the function identify to interactively detect the row number of individual
#genes by clicking on the plot. One can then recover the gene identifiers by saving the resulting indices:

idx1 <- identify(res1$baseMean, res1$log2FoldChange)
rownames(res1)[idx1]


#Histogram
hist(res1$pvalue, breaks=20, col="grey")
hist(res1$padj, breaks=20, col="grey")


#Plot counts
#Mature
plotCounts(dds1, gene=which.min(res1$padj), intgroup= c("Treatment"))


#For customized plotting, an argument returnData specifies that the function should only return a data.frame for plotting with ggplot.

#Mature
d1 <- plotCounts(dds1, gene=which.min(res1$padj), intgroup="Treatment", 
                returnData=TRUE)


#Mature ggplot

ggplot(d1, aes(x=Treatment, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + 
  scale_y_log10(breaks=c(200,2000, 10000)) #y-axis coordinate

#More information on results columns

mcols(res1)$description

```
#Volcano plot
```{r}
library(dplyr)
resa <- as.data.frame(res1)

# The significantly differentially expressed genes are the ones found in the upper-left and upper-right corners.
# Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2FoldChange respectively positive or negative)

# add a column of NAs
resa$diffexpressed <- "NO"

# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 

resa$diffexpressed[resa$log2FoldChange > 1 & resa$padj < 0.01] <- "UP"

# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
resa$diffexpressed[resa$log2FoldChange < -1 & resa$padj < 0.05] <- "DOWN"


# Re-plot but this time color the points with "diffexpressed"
p <- ggplot(resa, aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed)) + geom_point() + theme_minimal()

# Add lines as before...
p2 <- p + geom_vline(xintercept=c(-0.6, 0.6), col="red") +
        geom_hline(yintercept=-log10(0.05), col="red")

## Change point color 

# 1. by default, it is assigned to the categories in an alphabetical order):
p3 <- p2 + scale_color_manual(values=c("blue", "black", "red"))

# 2. to automate a bit: create a named vector: the values are the colors to be used, the names are the categories they will be assigned to:
mycolors <- c("blue", "red", "black")
names(mycolors) <- c("DOWN", "UP", "NO")
p3 <- p2 + scale_colour_manual(values = mycolors)

# Now write down the name of genes beside the points...
# Create a new column "delabel" to de, that will contain the name of genes differentially expressed (NA in case they are not)
resa$delabel <- NA
resa$miRNA = gsub("\\.","", row.names(resa))

resa$delabel[resa$diffexpressed != "NO"] <- resa$miRNA[resa$diffexpressed != "NO"]

ggplot(resa, aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed, label=delabel)) +
    geom_point() + 
    theme_minimal() +
    geom_text()

# Finally, we can organize the labels nicely using the "ggrepel" package and the geom_text_repel() function
# load library
library(ggrepel)
# plot adding up all layers we have seen so far
ggplot(resa, aes(x=log2FoldChange, y=-log10(padj), col=diffexpressed, label=delabel)) +
        geom_point() + 
        theme_minimal() +
        geom_text_repel() +
        scale_color_manual(values=c("blue", "black", "red")) +
        geom_vline(xintercept=c(-0.6, 0.6), col="red") +
        geom_hline(yintercept=-log10(0.05), col="red")
```
#To select the top miRNA
```{r}
resa = resa[order(abs(resa$log2FoldChange), decreasing = TRUE),]
topGene = rownames(resa)[1]
```

### Multi-factor designs?
```{r MultiFactor}
#Mature
kable(colData(dds1)) %>% kable_styling() 
ddsMF1 <- dds1
levels(ddsMF1$Parity)
levels(ddsMF1$Treatment)
levels(ddsMF1$Lactation_stage)

levels(ddsMF1$Parity) <- sub("-.*", "", levels(ddsMF1$Parity))
levels(ddsMF1$Parity)

levels(ddsMF1$Treatment) <- sub("-.*", "", levels(ddsMF1$Treatment))
levels(ddsMF1$Treatment)

levels(ddsMF1$Lactation_stage) <- sub("-.*", "", levels(ddsMF1$Lactation_stage))
levels(ddsMF1$Lactation_stage)

design(ddsMF1) <- formula(~ Lactation_stage + Treatment)
ddsMF1 <- DESeq(ddsMF1)

resMF1 <- results(ddsMF1)
kable(head(resMF1)) %>% kable_styling()

resMFType1a <- results(ddsMF1, contrast=c("Treatment", "S.chromogenes", "Control"))
kable(head(resMFType1a)) %>% kable_styling()

resMFType1b <- results(ddsMF1, contrast=c("Treatment", "S.chromogenes", "Control"))
kable(head(resMFType1b)) %>% kable_styling()


```

#Heat map
```{r}
#### Data quality assessment by sample clustering and visualization

#Mature
select1 <- order(rowMeans(counts(dds1,normalized=TRUE)), 
                decreasing=TRUE)[1:20]

df1 <- as.data.frame(colData(dds1)[, c("Lactation_stage", "Parity", "DIM", "Herd_name", "Udder_quarter", "Cow_number", "Treatment")]) 


rownames(df1) <- colnames(assay(ntd1)[select1,])
colnames(df1) <- c("Lactation_stage", "Parity", "DIM", "Herd_name", "Udder_quarter", "Cow_number", "Treatment")


pheat1 <- pheatmap(assay(ntd1)[select1,], 
                  cluster_rows=FALSE, 
                  show_rownames=FALSE,
                  cluster_cols=FALSE, 
                  annotation_col=df1)

pheat_v1 <- pheatmap(assay(vsd1)[select1,], cluster_rows=FALSE, show_rownames=FALSE,
                    cluster_cols=FALSE, annotation_col=df1)
```

#Heat map of sample to sample distances 
```{r}
#Mature
sampleDists1 <- dist(t(assay(vsd1)))
sampleDistMatrix1 <- as.matrix(sampleDists1)
rownames(sampleDistMatrix1) <- paste(vsd1$Treatment,colnames(vsd1),sep = "_")
colnames(sampleDistMatrix1) <- NULL
colors1 <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

phm1 <- pheatmap(sampleDistMatrix1,
         clustering_distance_rows=sampleDists1,
         clustering_distance_cols=sampleDists1,
         col=colors1)

#Mature
sampleDists3 <- dist(t(assay(rld1)))
sampleDistMatrix3 <- as.matrix(sampleDists3)
rownames(sampleDistMatrix3) <- paste(rld1$Treatment, rld1$Lactation_stage,sep = "_")
colnames(sampleDistMatrix3) <- NULL
colors3 <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

phm3 <- pheatmap(sampleDistMatrix3,
         clustering_distance_rows=sampleDists3,
         clustering_distance_cols=sampleDists3,
         col=colors3)
```
#Heatmap complete
```{r}
library(ggplot2)
resb <- as.data.frame(cts_snoRNA)
Data.log.scale <- scale(log10(resb +1))
y <- as.matrix(Data.log.scale)

hr <- hclust(as.dist(1 - cor(t(y), method = "pearson")), method = "average")
hc <- hclust(as.dist(1 - cor(y, method = "spearman")), method = "average")
mycl <-  cutree(hr, h = max(hr$height) / 1.01)
mycolhc <-   rainbow(length(unique(mycl)), start = 0.1, end = 0.9)
mycolhc <- mycolhc[as.vector(mycl)]
cluster <- as.matrix(mycl)

library(gplots)
heatmap.2(
  y,
  Rowv = as.dendrogram(hr),
  Colv = as.dendrogram(hc),
  col = greenred(75),
  density.info = "none",
  trace = "none",
  dendrogram = "both",
  scale = "row",
  labRow = NULL,
  labCol = NULL,
   margins = c(5, 10),
  RowSideColors = mycolhc)
```

#Principal component analysis
```{r}
#Mature
plotPCA(vsd1,intgroup=c("Treatment"))

#using the ggplot function to customize the PCA plot
#Mature
pcaData1 <- plotPCA(vsd1, intgroup=c("Treatment"), returnData=TRUE)

#intgroup=c("condition", "type")
#Mature
percentVar1 <- round(100 * attr(pcaData1, "percentVar1"))
pca1 <- ggplot(pcaData1, aes(PC1, PC2, color = Treatment)) + 
  geom_point(size = 3) + geom_text(aes(label = name), vjust = 1) + 
  xlab(paste0("PC1: ", percentVar1[1],"% variance")) + 
  ylab(paste0("PC2: ", percentVar1[2],"% variance")) + 
  coord_fixed()
pca1

#rld -PCA plot
#Mature
plotPCA(rld1, intgroup = c("Treatment", "Lactation_stage"))

(pcadata <- plotPCA(rld1, intgroup = c( "Treatment", "Lactation_stage"), returnData=TRUE))
```
# Variations to the standard workflow
```{r Variations}
#Mature
dds1 <- estimateSizeFactors(dds1)
dds1 <- estimateDispersions(dds1)
dds1 <- nbinomWaldTest(dds1)
head(resa)
```
###order results table by the smallest p value  

```{r}
resOrdered <- res1[order(res1$pvalue),]

summary(resOrdered)
##p-values and adjusted p-values

# order results table by the smallest p value:
sum(res1$pvalue < 0.05, na.rm = TRUE) 

sum(res1$padj < 0.05, na.rm=TRUE)
```
####Exporting only the results which pass an adjusted p value threshold
```{r}
resSig <- subset(resOrdered, padj < 0.05)

kable(head(resSig)) %>% kable_styling()
#Label

resSig$diffexpressed[resSig$log2FoldChange > 1 & resSig$padj < 0.05] <- "UP"

# if log2Foldchange < -1 and pvalue < 0.05, set as "DOWN"

resSig$diffexpressed[resSig$log2FoldChange < -1 & resSig$padj < 0.05] <- "DOWN"

resSig$miRNA = gsub("\\.","", row.names(resSig))
```

####merge with normalized count data
```{r}
resdata <- merge(as.data.frame(resSig), as.data.frame(vsd1@assays@data@listData[[1]]), by="row.names", sort=FALSE)

write.csv(as.data.frame(resdata), file="SC_mRNA1.csv")

```